---
title: "Re-predict GEDI L4A AGBD from stored `xvar` using the revised R003 EBT_SA model"
output:
  html_document:
    df_print: paged
---

This notebook applies the revised R003 'EBT_SA' biomass model to a Version 3 (V003) GEDI L4A granule for South American (SA) forest strata, without re-deriving the model's predictors (relative height, RH) from the Level L2A GEDI product.

There are two prediction strata that use the 'EBT_SA' model: 'EBT_SA' and 'DBT_SA'. The code below finds this from the granule's own ANCILLARY/model_data (looking for model_name == 'EBT_SA') rather than hardcoding the pair.

Each GEDI shot already stores its fully-built (offset-applied, transformed)
predictor design matrix in the `xvar` dataset, and the model actually applied
when the granule was produced is recorded per prediction stratum in
`ANCILLARY/model_data`. There is no need to go back to the L2A product for re-deriving 
the RH metrics and re-predicting AGBD with new model coefficients is just the following steps:

1. Find all footprints where `predict_stratum` was `'EBT_SA'` or `'DBT_SA'`
   when the granule was produced.
2. Confirm the predictor definition (`rh_index`, `predictor_id`,
   `x_transform`) for that stratum still matches the revised model —
   this is what makes reusing `xvar` valid instead of a coincidence.
3. Re-apply `agbd_t = [1, xvar] %*% par_new`, back-transform, and re-apply the
   bias correction.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Prerequisites

- **Granule file**: this notebook expects the L4A `.h5` granule in the same directory. Adjust 
  the path if yours lives elsewhere.
- **Package**: Requires the Bioconductor `rhdf5` package (`BiocManager::install('rhdf5')`).
Unlike some HDF5 R bindings, `rhdf5` links its own bundled HDF5 build (via the
`rhdf5lib` package) rather than whatever HDF5 happens to be installed on the
system, so there's nothing else to configure.

```{r libs}
if (!requireNamespace("rhdf5", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("rhdf5")
}

library(rhdf5)

NODATA <- -9999.0

# A V003 L4A granule (Amazon basin overpass) with a large EBT_SA population
GRANULE_FILE <- 'GEDI04_A_2022314110220_O22155_04_T02087_02_004_01_V003.h5'

# The model_name (== fit_stratum bucket in the selection JSON) whose coefficients
# were revised on 2026-07-22
CHANGED_MODEL_NAME <- 'EBT_SA'

# NEW_MODEL holds the revised coefficients from issgedi_l4a_r003_estimators_20260722.json
# group 4 / EBT_SA / selection 3 (identical to selection 4, used by DBT_SA --
# both prediction strata now share this exact fit)
NEW_MODEL <- list(
  rh_index = c(50L, 98L),
  predictor_id = c(1L, 2L),
  x_transform = 'sqrt',
  y_transform = 'sqrt',
  bias_correction_name = 'Snowdon',
  bias_correction_value = 1.15323996418645,
  par = c(-133.884363713764, 6.95698142185319, 6.28656682203761),
  vcov = c(2.50305363959644, -0.111702554712149, -0.114315207874669,
           -0.111702554712149, 0.0654051766999497, -0.0523471595323726,
           -0.114315207874669, -0.0523471595323726, 0.0598715899876015),
  rse = 3.53298787494478,
  dof = 3485,
  response_max_value = 1578.0,
  predictor_max_value = c(12.2270192606375, 12.7730184373154)
)
```

```{r functions}
# The EBT_SA model is a square-root model with a Snowdon bias correction --
# hardcoded rather than dispatched on NEW_MODEL$y_transform/bias_correction_name,
# since this is the only model this notebook ever applies.
back_transform <- function(agbd_t) {
  agbd_t^2
}

apply_bias_correction <- function(agbd, value) {
  agbd * value
}

# Per-stratum model actually used to produce this granule -- the model_name it
# was routed to, plus its predictor definition (rh_index / predictor_id /
# x_transform) -- read from ANCILLARY/model_data.
#
# compoundAsDataFrame = FALSE is required here: rhdf5's default (TRUE) can't
# coerce a compound field of type ARRAY (rh_index, predictor_id, par, vcov, ...)
# into a data.frame column and silently replaces those fields with NA. With
# FALSE, every row is read in one call and array fields come back as proper
# (field_length x n_rows) matrices -- no truncation, no row-by-row workaround.
get_baked_in_models <- function(fid) {
  md <- h5read(fid, "ANCILLARY/model_data", compoundAsDataFrame = FALSE)
  models <- list()
  for (i in seq_along(md$predict_stratum)) {
    npar <- as.integer(md$npar[i])
    models[[md$predict_stratum[i]]] <- list(
      model_name = md$model_name[i],
      rh_index = as.integer(md$rh_index[1:(npar - 1), i]),
      predictor_id = as.integer(md$predictor_id[1:(npar - 1), i]),
      x_transform = md$x_transform[i]
    )
  }
  models
}

# Prediction strata routed through the revised model, restricted to those
# whose predictor definition is unchanged from what's baked into the granule.
get_compatible_strata <- function(baked_in_models) {
  compatible <- character(0)
  for (stratum in names(baked_in_models)) {
    baked_in <- baked_in_models[[stratum]]
    if (baked_in$model_name != CHANGED_MODEL_NAME) next
    ok <- identical(baked_in$rh_index, as.integer(NEW_MODEL$rh_index)) &&
      identical(baked_in$predictor_id, as.integer(NEW_MODEL$predictor_id)) &&
      identical(baked_in$x_transform, NEW_MODEL$x_transform)
    if (ok) {
      compatible <- c(compatible, stratum)
    } else {
      cat(sprintf('  SKIPPED %s: predictor definition changed, xvar reuse invalid\n', stratum))
    }
  }
  compatible
}

# Re-predict AGBD from a stored predictor matrix using the revised model.
#
# `xvar` is (npar_max - 1) x nshots: rhdf5, like other HDF5 R bindings,
# reverses HDF5 dataset dimension order relative to h5py/numpy (HDF5 is
# row-major, R is column-major), so a granule's (nshots, npar_max - 1)
# dataset arrives here transposed.
recompute_agbd_from_xvar <- function(xvar, predict_stratum, compatible_strata) {
  agbd_new <- rep(NODATA, length(predict_stratum))
  par <- NEW_MODEL$par
  npar <- length(par)

  for (stratum in compatible_strata) {
    idx <- which(predict_stratum == stratum)
    if (length(idx) == 0) next
    xvar_s <- t(xvar[1:(npar - 1), idx, drop = FALSE])  # nidx x (npar - 1)
    valid <- rowSums(xvar_s == NODATA) == 0
    if (!any(valid)) next

    predictors <- cbind(1, xvar_s[valid, , drop = FALSE])
    agbd_t <- as.vector(predictors %*% par)
    agbd <- back_transform(agbd_t)
    agbd <- apply_bias_correction(agbd, NEW_MODEL$bias_correction_value)
    agbd_new[idx[valid]] <- agbd
  }

  agbd_new
}
```

## Read the granule and re-predict AGBD per beam

Only `xvar`, `predict_stratum`, `agbd` (old estimate), `l4a_quality_flag_rel3`,
`lat_lowestmode`/`lon_lowestmode`, and `ANCILLARY/model_data` are needed.

```{r read-granule}
fid <- H5Fopen(GRANULE_FILE)
beams <- h5ls(fid, recursive = FALSE)$name
beams <- beams[startsWith(beams, "BEAM")]
baked_in_models <- get_baked_in_models(fid)

cat(sprintf("Finding strata routed through the '%s' model:\n", CHANGED_MODEL_NAME))
compatible_strata <- get_compatible_strata(baked_in_models)

agbd_old_list <- list(); agbd_new_list <- list(); quality_list <- list(); stratum_list <- list()
lon_list <- list(); lat_list <- list()

for (beam in beams) {
  xvar <- h5read(fid, paste0(beam, "/xvar"))
  predict_stratum <- h5read(fid, paste0(beam, "/predict_stratum"))
  agbd_old <- h5read(fid, paste0(beam, "/agbd"))
  quality_flag <- h5read(fid, paste0(beam, "/l4a_quality_flag_rel3"))
  lon <- h5read(fid, paste0(beam, "/lon_lowestmode"))
  lat <- h5read(fid, paste0(beam, "/lat_lowestmode"))

  agbd_new <- recompute_agbd_from_xvar(xvar, predict_stratum, compatible_strata)

  agbd_old_list[[beam]] <- agbd_old
  agbd_new_list[[beam]] <- agbd_new
  quality_list[[beam]] <- quality_flag
  stratum_list[[beam]] <- predict_stratum
  lon_list[[beam]] <- lon
  lat_list[[beam]] <- lat
}
H5Fclose(fid)

agbd_old <- unlist(agbd_old_list, use.names = FALSE)
agbd_new <- unlist(agbd_new_list, use.names = FALSE)
quality_flag <- unlist(quality_list, use.names = FALSE)
predict_stratum <- unlist(stratum_list, use.names = FALSE)
lon <- unlist(lon_list, use.names = FALSE)
lat <- unlist(lat_list, use.names = FALSE)

valid_coord <- (lon > -180) & (lon < 180)
cat(sprintf('Granule: %s\n', basename(GRANULE_FILE)))
cat(sprintf('Bounding box: lon [%.2f, %.2f], lat [%.2f, %.2f]  (South America)\n',
            min(lon[valid_coord]), max(lon[valid_coord]),
            min(lat[valid_coord]), max(lat[valid_coord])))
cat(sprintf('Total shots: %d\n', length(agbd_old)))
cat(sprintf('Strata updated: %s\n', paste(sort(compatible_strata), collapse = ', ')))
```

## Effect of the revised estimators, by prediction stratum

Restricted to `l4a_quality_flag_rel3 == 1` shots (highest-quality shots).

```{r summary-table}
quality <- (quality_flag == 1) & (agbd_old != NODATA) & (agbd_new != NODATA)
cat(sprintf('Quality shots: %d / %d\n\n', sum(quality), length(agbd_old)))
cat(sprintf('%-10s %8s %12s %12s %12s\n', 'stratum', 'n', 'mean_before', 'mean_after', 'mean_delta'))
for (stratum in sort(unique(predict_stratum[quality]))) {
  m <- quality & (predict_stratum == stratum)
  delta <- agbd_new[m] - agbd_old[m]
  cat(sprintf('%-10s %8d %12.2f %12.2f %12.3f\n',
              stratum, sum(m), mean(agbd_old[m]), mean(agbd_new[m]), mean(delta)))
}

cat(sprintf('\nOverall mean AGBD before: %.2f Mg/ha\n', mean(agbd_old[quality])))
cat(sprintf('Overall mean AGBD after:  %.2f Mg/ha\n', mean(agbd_new[quality])))
```

### Distribution of the per-shot change

The histogram shows the per-shot AGBD change (`after - before`). The dashed line
marks the mean. A distribution sitting mostly left of zero confirms the revision
reduces biomass estimates across the quality-flagged shots, not just on average.

```{r delta-histogram, fig.width=6.5, fig.height=4.5}
COLOR_DELTA <- '#eb6834'  # categorical slot 2 (orange)

before <- agbd_old[quality]
after <- agbd_new[quality]
delta <- after - before
delta_mean <- mean(delta)
delta_sd <- sd(delta)

hist(delta, breaks = 60, col = COLOR_DELTA, border = 'white',
     xlab = 'AGBD after - before (Mg/ha)', ylab = 'Shot count',
     main = sprintf('EBT_SA + DBT_SA quality-flagged shot AGBD change (n=%d)', sum(quality)))
abline(v = delta_mean, col = 'grey20', lwd = 1.5, lty = 2)
legend('topright', bty = 'n',
       legend = sprintf('mean = %.2f Mg/ha\ns.d. = %.2f Mg/ha', delta_mean, delta_sd))
```
### Where in the distribution is the change?

The Q-Q plot compares the before/after AGBD distributions percentile-by-
percentile against the 1:1 line. Points falling below the line mean the revised
estimate is lower at that percentile; the growing gap toward the high end shows
the reduction is largest for the highest-biomass shots.

```{r ebt-sa-qq, fig.width=5.5, fig.height=5.5}
percentiles <- 1:99 / 100
before_q <- quantile(before, percentiles)
after_q <- quantile(after, percentiles)
lim <- max(before_q, after_q, 600)

plot(before_q, after_q, pch = 21, col = COLOR_DELTA, bg = NA, lwd = 1.2, cex = 1.1,
     xlim = c(0, lim), ylim = c(0, lim), asp = 1,
     xlab = 'AGBD before, percentile (Mg/ha)', ylab = 'AGBD after, percentile (Mg/ha)',
     main = sprintf('EBT_SA + DBT_SA quality-flagged shot Q-Q plot (n=%d)', sum(quality)))
abline(0, 1, col = 'grey30', lwd = 1, lty = 2)
legend('topleft', legend = '1:1', col = 'grey30', lwd = 1, lty = 2, bty = 'n')
```
